### 2017-10-05 response/response-jde-1.tex

[RESPONSE:] We have changed the equation and the associated text.

\item What other time-varying covariates are available in the survey? It
would be interesting to consider the extent to which the crop loss
results are explained by other household-level idiosyncratic variation.

[RESPONSE:]

There are three time-varying covariates in the survey that could 
potentially affect crop loss or the outcome variables directly:
rainfall, prices, and loss of livestock.
Other time-varying variables are illness and health variables, which 
we already show that these are unlikely to explain the results we find.

@@

[also speaks to whether this is a conscious decision or not] 
The table also supports our conjecture that the estimated effects 
[might need a less causal word] 
of crop loss capture a conscious decision. 
If shocks lead to lower probability of pregnancy and birth because
of other factors than the decision to use contraceptive we should see
a negative and statistically significant coefficient for the crop loss
dummy by itself.
We do not.
Instead, what we get is a substantial negative coefficient for the interaction
between contraceptive use and crop loss and a coefficient for crop loss 
on its own that is positive and far from statistically significant.

\item Several appendix tables interact crop loss with initial assets in
Round 1 of the survey. This exercise is interesting and worthy of doing,
given that one might expect asset-poor households to have particular
difficulty in smoothing consumption. But the authors should leave Round
1 outcomes out of this regression.

[RESPONSE:] We have added the results where round 1 has been excluded
to the relevant tables.
The results are broadly consistent with what we originally found, 
although less likely to be statistically significant 
because of the lower variation in crop loss and the smaller sample
size.

@@

Neighbors may be affected an individuals crop loss. For one, price of
food goes up. This could be a positive income shock for families with
crops, but bad for families working in other areas.

Suggestion: Aggregate crop loss to the community level and use that as
the treatment variable.

[RESPONSE:]

We already include a version of this in the paper.
Table A-17 TK include the fraction of households in the community 
who have experienced a crop (excluding the household itself).
The sign of the village level crop loss are generally the same as
those for the household crop loss, although---with the exception
of traditional contraceptives---the village level variable is not
statistically significant.
Furthermore, the crop loss coefficients decline only slightly
when including village level crop loss indicating that there is
only relatively mild spill-over between households for this
particular type of shock.

### 2017-10-07 response/response-jde-1.tex

% This interesting paper analyzes the relationship between income shocks
% (as proxied by self-reported crop loss) and marital fertility, using a
% small Tanzanian panel survey from the early 1990s. A key feature of the
% paper is that the authors are able to observe use of both traditional
% and modern birth control techniques. In regression specifications with
% and without community or woman fixed effects, the authors find that crop
% loss is associated with decreases in the probability of conceiving, with
% a corresponding increase in use of traditional birth control techniques
% but not modern contraceptives.

The paper could benefit from focusing its motivation somewhat more. An
important point (that the authors do not emphasize enough) is that use
of modern contraception in the sample is extremely low, so the lack of a
relationship between shocks and use of modern methods is not surprising.
The more noteworthy result is that use of traditional methods responds
to shocks. In this sense, the paper makes an important contribution to
understanding how populations control fertility before widespread
adoption of modern contraception. For instance, it is well known that
marital fertility in Europe declined largely before modern contraception
became available, but the methods couples used to achieve this reduction
are not well documented (see, for example, Guinnane JEL 2011). The paper
would be even more compelling if the authors focused the paper's
motivation on the questions of whether and how couples modify fertility
to respond to changing economic conditions before the advent of
modern contraception. That motivation would also help justify their
choice of dataset, which otherwise seems somewhat small and out of date
relative to other available panel datasets from developing countries.

[RESPONSE:]

We have edited the introduction to highlight the high use of traditional 
contraception and that most of the increase in contraception after a 
shock comes from traditional methods.
We have also added an additional point to the discussion of the
main contributions of the paper.
In addition to the \citet{Guinnane2011} paper, we also cite a just
published paper on spacing in pre-fertility transition England
\citep{Cinnirella2017} and \citet{Michael1976}, \citet{David1986},
and \citet{Santow1995}
on the role of different types of traditional contraception in
controlling fertility before the emergence of modern contraception.

In the Conclusion we already discussed that our results were
especially of interest because of the composition of contraceptive
methods used.
We have edited the discussion to further clarify the major role
of traditional contraceptives.
In addition, we have recast the second to last paragraph to
highlight the connection to long-term fertility and the literature
on fertility transition in Europe and the US.

% [should the footnote on historical control of fertility be moved up
% to why this is important?]

% [few data sets with panel data on contraceptive use and economic
% circumstances]

@@

[TK summing up]

% Reduction in pregnancy is 0.3 for abstinence with crop loss
% Reduction in pregnancy is 0.2 for modern contraceptives with crop loss

\item The authors analyze the fertility of women who were married at the
start of the sample period. As a separate analysis, it would be
interesting to know whether marriage hazards change with crop loss in
the parents' home, although perhaps the authors do not have the sample
size to study this question.

[RESPONSE:] That is, indeed, a very interesting question, but we have 
decided not to pursue it at the moment for three reasons.
First, there are already two papers that examine this question, 
albeit for other countries \citet{mbiti08,Hoogeveen2011}.
Second, as those two papers show, the length of our paper would have
to increase substantially to do this question justice, and the Editor
has asked us to not increase the length of the paper.
Finally, the sample size is, indeed, on the small size to really get 
at this question.

### 2017-10-07 paper/incomeShocks-jde-r1.tex
Remove map of Kagera
Respond to R1 on historical fertility and traditional contraceptives

How shocks affect household decisions on fertility in developing countries has 
received little attention in the literature.
There is evidence that shocks can affect the timing of births, but it is unclear
whether the cause is households' intentional planning or unintended consequences 
of the shocks \citep{pitt98b,lindstrom99,Evans2010,Portner2014}.%
\footnote{
There is also a related literature on the effects of weather, especially temperature,
on timing of fertility \citep[see, for example,][]{Barreca2015}.
%  and TK Josh paper.
In a different vein, a monthlong blackout in Zanzibar increased 
births 8 to 10 months later \citep{Burlando2014}.
There is also a large literature examining how households in developing countries
cope with shocks in areas such as labor supply, crop choice, and education.
} 
Furthermore, standard models of fertility focus mainly on the effects of time
invariant household and individual characteristics on fertility outcomes, and 
tend to ignore the dynamic aspects of fertility decisions.%
\footnote{
For an early exception see \cite{Newman1984}.
}

We seek to answer two key questions in this paper.
First, do income shocks, measured by accidental crop loss in Tanzania, affect timing of 
fertility?
Second, do changes in fertility timing, if any, arise from an intentional decision
to postpone fertility?

There are three reasons why a household may want to delay fertility in the
event of a shock.
First, children are costly in the short run.
Having more children may eventually contribute to the household's production 
and help it overcome shocks, but the short-term impact on availability 
of resources is almost certainly negative \citep{Portner2014}.
Secondly, diverting time away from market work to child care will be even more costly
if households need to respond to a shock by working more \citep{kochar99};
although it is also possible that a shock will temporarily lower women's cost 
of time, making it more attractive to have a birth now.
Finally, the household may realize that children born following a shock 
are more likely to be malnourished and have worse health outcomes, and therefore 
decide to postpone having the next child \citep{Portner2010}.

Observing a decline in pregnancies and birth rates following a shock is, however, 
not evidence of a conscious decision to change the timing of fertility.
A shock may lead one or more members of the household to migrate 
in search of better economic opportunities.
If either the husband or the wife is gone for extended periods, this 
will lower the probability of conception.
Furthermore, a shock may reduce intercourse frequency---either
from psychological effects or from increased hours of work.
We would also observe a decline in births if people delay marriage or if there
is an increase in the dissolution of marriage.
Finally, severe shocks can lead to starvation and health problems \citep{lindstrom99}.
Malnourished pregnant mothers are more likely to have stillbirths,%
\footnote{
See \cite{Hernandez-Julian2014} for a review of the literature and an 
analysis of the effect of famine on stillbirths.
}
and non-pregnant women may see an increased incidence of secondary sterility
from a reduction in age of menopause or famine induced amenorrhea.

Understanding the relation between shocks and fertility decisions is important for
three reasons.
First, as many households in developing countries experience a large number of shocks, 
it is important to understand the factors that help or hinder households' recovery 
after a shock.
Second, children born immediately after a shock fare worse than other children do.
In the short- and medium-term, they have worse health as measured by height-for-age 
Z-scores \citep{Portner2010}, and
long-term effects include worse self-reported health, lower schooling, and less
wealth \citep{Maccini2009}.
% This could be better explained
Finally, parents do not directly control fertility and its timing, but rather 
control type and intensiveness of contraceptive use and the frequency and timing of 
sexual activity.
Hence, it helps us understand if, and how, households regulate their fertility.
This, in turn, can inform policies on availability and targeting
of family planning services.

Most of the prior literature on the impact of economic and other shocks on fertility has
used historic data on what are now developed countries.
Data from Rouen, France, over the period 1681-1787, show that increases in the
price of wheat led to a dramatic fertility decline for the urban poor, 
but fertility of the urban wealthy was unaffected \citep{Galloway1987}.
Similarly, English data from 1542 to 1800 show that mortality shocks led to short-term
fertility declines, with the largest decline typically occurring  the year following 
the shock \citep{Bailey1998}.%
\footnote{
This data also indicated that an increase in real wage led to an increase in short-term 
fertility.
Increases in wage rate and declines in child mortality explain a significant part of the 
long-term fertility decline in Sweden during the period 1736 to 1946 \citep{Eckstein1999a}.
In another study on Sweden, increases in the value of women's time relative to men's 
time---arising from increases in the price of butter---led to a decline in fertility during 
the  1860-1910 period \citep{Schultz1985}. 
}

In pre-industrial Sweden, a positive crop shock, positive weather shock, and positive wage 
shock increased fertility for a five- to ten-year period following the shocks \citep{Eckstein1985}.
However, the increase in birth rate changes only the timing of births and has no cumulative 
effect on long-term fertility rate.
Similarly, the increasing price of rye in Southern Sweden led landless and semi-landless 
families to reduce their fertility, whereas landed farmers showed no effect 
\citep{bengtsson06}.

@@

Many studies examine how a variety of factors affect contraception use, such 
as, 
schooling, 
focused information campaigns, and
participation in savings or credit group 
\citep{Ainsworth1996,Feyisetan1996,Steele2001,Chen2003}.
To our knowledge, only one study attempts to show the effect of shocks on 
contraception use \citep{Hernandez-Correa2010}. 
Cross-sectional data from Madagascar show that households that experience adverse events 
are more likely to use contraception.\footnote{%
Adverse events include economic and environmental aspect, such as: rise in input 
cost, rise in cost of goods, difficulty finding buyers, difficulty finding inputs, 
floods, late rains, early rains, droughts, pest problems, etc. 
} 
However, households may be able to anticipate many of these events before they 
occur and adjust their behavior accordingly.
Combined with possible correlation between the probability of
experiencing an adverse event and unobserved individual and household
characteristics, this may bias the estimates, and the cross-sectional
data allows for neither household nor individual level fixed effects
estimates.

We use longitudinal data from the Kagera region in Tanzania with information on accidental 
crop loss, fertility, and contraceptive use to estimate individual level fixed 
effects models of the effects of income shocks on both timing of fertility and 
contraceptive use.
We show that income shocks decrease the likelihood of pregnancies and births.
At the same time, shocks significantly increase use of contraceptives, 
especially traditional contraceptives.
We argue that the postponement of fertility in households experiencing income 
shocks is the result of the increased contraception use rather than a biological 
response or separation of spouses.

The paper makes two main contributions to the literature. 
First, this is the first paper to show that households respond to income shocks 
through family planning, which, in turn, affects the timing of child bearing. 
Our study provides evidence that the postponement of fertility occurs through a 
planned decision process rather than as an unplanned consequence.
Secondly, the paper contributes to the consumption smoothing literature as households 
affected by income shock use family planning as a mechanism to smooth their consumption. 
In the short run, bearing a child means diverting scarce resources away from other
useful purposes to the birth and maintenance of the child. 
Farmers can, therefore, partly smooth their consumption by delaying births during 
times of income shock.

@@

We use the Kagera Health and Development Survey (KHDS) conducted by the World Bank 
and the University of Dar es Salaam in the Kagera region in Tanzania. 
The survey covers four rounds from 1991 through 1994 and over 800 households, 
drawn from 51 communities (49 villages) in the six districts of Kagera.%
\footnote{
The sample selection was based on a variable probability sampling procedure 
(a two-stage, randomized stratified procedure) based on expected mortality.
In the first stage, based on the 1988 Tanzanian census, the census clusters were randomly 
selected after stratifying them according to mortality rates and agro-climatic zones. 
Households were then divided into ``high-risk'' and ``low-risk'' groups in the 
second stage, based on illness and death of households in the 12 months 
before the enumeration process.
Finally, households were randomly sampled from the groups.
For further details on the sample selection, please see World Bank (2004).
}
The average interval between each of the survey round was between six and seven months, 
and interviews of households took place on a rolling basis throughout the year. 
The Kagera region is in the northwestern corner of Tanzania, west of Lake Victoria, and 
borders Uganda, Rwanda, and Burundi as shown in Figure \ref{fig:kagera_map}. 
The population of the area---1.3 million in 1988---is 
overwhelmingly rural and primarily engaged in agriculture.

\begin{figure}[ht] 
 \centering
 \includegraphics[width=0.6\textwidth]{../staticPDF/kagera.pdf}
 \caption{Map of Kagera region}
  \label{fig:kagera_map}
\end{figure}

@@

Table \ref{tab:desc_stat_croploss} shows the main explanatory variable of interest 
and the 5 outcomes that we examine.
The main explanatory variable is a dummy for whether the household experienced 
an accidental loss of crops, which covers any crop lost due to insects, 
rodents, fire, rotting, or other calamities, greater than TZS 200 since the 
prior survey round (or prior 1 year for the first round).
% We create a dummy for accidental crop losses greater than 200 TZS,
% which is our preferred measure of income shock.
The TZS 200 cut-off is equivalent to approximately 5th 
percentile of all household savings.%
% That period is approximately 6 to 7 months for the last three surveys, but covers 
% the previous 12 months for the first survey. 
% [MAY HAVE TO EXPLAIN MORE ABOUT WHAT THIS MEANS FOR OUR ESTIMATES]
There is substantial variation in the number of households that experience a 
crop loss between the 4 survey rounds.
In the 12 months prior to the first survey round just over 70\% of households 
lost crops, whereas just over 25\% lost crops between the 1st and 2nd survey round.
That falls even further for the 3rd and 4th rounds, with less than 5\% 
reporting crop loss of more than TZS 200 for both.
Our results are robust to defining crop loss as any crop loss, with the 
coefficients slightly smaller and slightly less statistically 
significant as expected. 
% Crop losses between the valuation of 0-200 TZS are not coded as an income shock,
% because of the small magnitude (equal to  of savings). 
% and unlikely to affect fertility decisions of households. 
Furthermore, our results are also qualitatively similar if we use a log of the amount of 
crop lost or a linear specification of crop loss instead of the dummy variable.%
\footnote{
Results using these alternative specifications of crop loss are available in the Appendix.
}

@@

We estimate the relationship between household income shocks and decisions
on timing of fertility. 
To account for potential correlation between unobservable household
characteristics and both shocks and fertility decisions, we employ a fixed effect model.
Our estimates show that both the likelihood of pregnancy and birth decrease in
response to a crop loss and that contraceptive use, especially use of traditional 
methods like abstinence and rhythm, significantly increases.

@@

Based on this, our main conclusion is that households consciously delay 
fertility through adjustment in contraceptive use to cope 
with income shocks. 
This is especially of interest since there is little use of modern
contraceptives in this setting.
If traditional contraceptives have low efficacy, a natural question is 
how the substantial reductions in likelihood of births and pregnancy 
come about.
We present suggestive evidence that show that only those households
with a strong incentive, i.e. those exposed to crop loss, are 
effective users of traditional contraceptives.%
\footnote{
This is in line with the results in \cite{Rosenzweig1989},
where women with the highest opportunity cost of children
are also the most effective user of traditional contraceptives,
although the setting is very different.
}

@@

The data unfortunately do not allow us to establish whether there 
are long-term impacts on fertility outcomes.
Results from Guatemala indicate that only shocks that occur 
toward the end of a woman's reproductive life is likely to have an 
impact on final fertility \citep{Portner2014}.
There is, therefore, little reason to believe that temporary shocks,
like the crop losses studied here, lead to over-all changes in fertility.

Hence, an important unanswered question is what the implications
are for our understanding of people's ability to control long-term
fertility when there is limited access to modern contraceptives.
As we have shown, it is possible for households to, at least temporarily, 
limit fertility using traditional methods once the incentives are
strong enough.
What we do not know is whether the psychic cost of using traditional 
contraceptives is so high that although households are willing to use 
them to deal with short-term shocks, they become difficult to use for 
long-run moderation of fertility.
That studies of the introduction of modern family planning, using individual
level data, rarely find reductions in fertility larger than one child
is an indication that households were already able to limit long-term 
fertility using traditional contraceptives.%
\footnote{
See, for example, \cite{Portner2014a} and the literature surveyed therein.
}

This does not mean that the ability to postpone birth is unimportant
or that there is not an important role for modern contraceptives.
Children born after shocks are likely to fare worse than children born 
during better times.
The mother is also likely to fare better if she does not give birth 
immediately following an economic shock that may impact her ability 
to recover and care for the new child.
Our results imply that providing better access to contraceptives 
during times of economic shocks could improve the ability of households 
to postpone births.
In turn, this could lead to better mother and child health outcomes.

>>>>

How shocks affect household decisions on fertility in developing countries has 
received little attention in the literature.
A correlation between shocks and the timing of births is well-established, but it is 
unclear whether the cause is households' intentional planning or unintended consequences 
of the shocks \citep{pitt98b,lindstrom99,Evans2010,Portner2014}.%
\footnote{
There is a related literature on the effects of weather, especially temperature,
on timing of fertility \citep[see, for example,][]{Barreca2015}.
%  and TK Josh paper.
In a different vein, a monthlong blackout in Zanzibar increased 
births 8 to 10 months later \citep{Burlando2014}.
There is also a large literature examining how households in developing countries
cope with shocks in areas such as labor supply, crop choice, and education.
}
There is, in general, little evidence on what determines the timing 
of fertility in developing countries, with most of the literature instead 
focusing on the number of children born \citep{Portner2018}.
% [Little evidence in general on timing of fertility in developing countries;
% most of the research has focused on completed fertility / children-ever-born] 
% Furthermore, standard models of fertility focus mainly on the effects of time
% invariant household and individual characteristics on fertility outcomes, and 
% tend to ignore the dynamic aspects of fertility decisions.%
% \footnote{
% For an early exception see \cite{Newman1984}.
% }

We seek to answer two key questions in this paper.
First, do income shocks, measured by accidental crop loss in Tanzania, affect the
timing of fertility?
Second, do changes in fertility timing, if any, arise from an intentional decision
to postpone fertility?

There are three reasons a household may want to delay fertility in the
event of a shock.
First, children are costly in the short run.
Having more children may eventually contribute to the household's production 
and help it overcome shocks, but the short-term impact on availability 
of resources is almost certainly negative \citep{Portner2014}.
Secondly, diverting time away from market work to child care will be even more 
costly if the household needs to respond to a shock by working more \citep{kochar99}.%
\footnote{
It is, however, also possible that a shock could temporarily lower women's 
cost of time, making it more attractive to have a birth now.
}
Finally, the household may realize that children born following a shock 
are more likely to be malnourished and have worse health outcomes, and therefore 
decide to postpone having the next child \citep{Portner2010}.

Observing a decline in pregnancies and birth rates following a shock is, however, 
not evidence of a conscious decision to change the timing of fertility.
A shock may, for example, lead one or more members of the household to 
migrate in search of better economic opportunities, and if either the husband 
or the wife are gone for extended periods, this will lower the probability of 
conception.
Furthermore, a shock may reduce intercourse frequency---either for
physiological or psychological reasons or from increased hours of work.
We would also observe a decline in births if people delay marriage or 
if there is an increase in the dissolution of marriage.
Finally, severe shocks can lead to starvation and health problems \citep{lindstrom99}.
Malnourished pregnant mothers are more likely to have stillbirths,
and non-pregnant women may see an increased incidence of secondary sterility
from a reduction in age of menopause or famine induced amenorrhea.%
\footnote{
See \cite{Hernandez-Julian2014} for a review of the literature and an 
analysis of the effect of famine on stillbirths.
}

Understanding the relation between shocks and fertility decisions is important for
three reasons.
First, as many households in developing countries experience a large number of shocks, 
it is important to understand the factors that help or hinder households' recovery 
after a shock.
Second, children born immediately after a shock fare worse than other children do.
In the short- and medium-term, they have worse health as measured by height-for-age 
Z-scores \citep{Portner2010}, and
long-term effects include worse self-reported health, lower schooling, and less
wealth \citep{Maccini2009}.
% This could be better explained
Finally, it helps us understand if, and how, households regulate their fertility.
% parents do not directly control fertility and its timing, but rather 
% control type and intensiveness of contraceptive use and the frequency and timing of 
% sexual activity.
This, in turn, can inform policies on availability and targeting
of family planning services.

Most of the prior literature on the impact of economic and other shocks on fertility 
has used historic data on what are now developed countries.
Data from Rouen, France, over the period 1681-1787, show that increases in the
price of wheat led to a dramatic fertility decline for the urban poor, 
whereas fertility of the urban wealthy was unaffected \citep{Galloway1987}.
Similarly, English data from 1542 to 1800 show that mortality shocks led to short-term
fertility declines, with the largest decline typically occurring  the year following 
the shock \citep{Bailey1998}.%
\footnote{
This data also indicated that an increase in real wage led to an increase in short-term 
fertility.
Increases in wage rate and declines in child mortality explain a significant part of the 
long-term fertility decline in Sweden during the period 1736 to 1946 \citep{Eckstein1999a}.
In another study on Sweden, increases in the value of women's time relative to men's 
time---arising from increases in the price of butter---led to a decline in fertility during 
the  1860-1910 period \citep{Schultz1985}. 
}
In pre-industrial Sweden, positive crop shock, positive weather shock, and 
positive wage shock all increased fertility for a five- to ten-year period following 
the shocks \citep{Eckstein1985}.
However, the increase in birth rate changed only the timing of births with no 
cumulative effect on the long-term fertility rate.
Similarly, the increasing price of rye in Southern Sweden led landless and 
semi-landless families to reduce their fertility, whereas landed farmers 
showed no change \citep{bengtsson06}.

@@

Many studies show how a variety of factors affect contraception use, such 
as, 
schooling, 
focused information campaigns, and
participation in savings or credit group 
\citep{Ainsworth1996,Feyisetan1996,Steele2001,Chen2003}.
To our knowledge, only one study attempts to show the effect of shocks on 
contraception use \citep{Hernandez-Correa2010}. 
Cross-sectional data from Madagascar show that households that experience adverse 
events are more likely to use contraception.\footnote{%
Adverse events include economic and environmental aspect, such as: rise in input 
cost, rise in cost of goods, difficulty finding buyers, difficulty finding inputs, 
floods, late rains, early rains, droughts, pest problems, etc. 
} 
However, 
% households may be able to anticipate many of these events before 
% they occur and adjust their behavior accordingly.
% Furthermore, 
the possible correlation between the probability of experiencing 
an adverse event and unobserved individual and household characteristics may 
bias the estimates, and the cross-sectional data allows for neither household 
nor individual level fixed effects estimates.

We use longitudinal data from the Kagera region in Tanzania with information on 
accidental crop loss, fertility, and contraceptive use to estimate individual 
level fixed effects models of the effects of income shocks on both timing of 
fertility and contraceptive use.
We show that income shocks decrease the likelihood of pregnancies and births.
At the same time, shocks significantly increase use of contraceptives.
This increase mainly occurs through the use of traditional contraceptives, 
which are the dominant form of contraception in the area.
We argue that the postponement of fertility in households experiencing income 
shocks is the result of the increased contraception use rather than a biological 
response or separation of spouses.

The paper makes three main contributions to the literature. 
First, this is the first paper to show that households respond to income shocks 
through contraceptive use, which, in turn, affects the timing of child bearing. 
% Our study provides evidence that the postponement of fertility occurs through a 
% planned decision process rather than as an unplanned consequence.
Second, with low rates of modern contraceptive use in the area, the paper contributes 
to the literature on how populations control fertility before widespread adoption of
modern contraception.%
\footnote{
There has long been a discussion of when and how families 
undertook fertility control before the emergence of modern 
contraceptives \citep{Guinnane2011,Cinnirella2017}.
Simulations show that it clearly is possible to substantially limit fertility
even with only access to traditional contraceptives, but there is only 
limited evidence available on exactly which methods were used 
\citep{Michael1976,David1986,Santow1995}.
}
Finally, the paper contributes to the consumption smoothing literature, as households 
affected by income shock use family planning as a mechanism to smooth their consumption. 
In the short run, bearing a child means diverting scarce resources away from other
useful purposes to the birth and maintenance of the child. 
Farmers can, therefore, partly smooth their consumption by delaying births during 
times of income shock.

@@

We use the Kagera Health and Development Survey (KHDS) conducted by the World Bank 
and the University of Dar es Salaam in the Kagera region in Tanzania. 
The survey covers four rounds from 1991 through 1994 and over 800 households, 
drawn from 51 communities (49 villages) in the six districts of Kagera.%
\footnote{
The sample selection was based on a variable probability sampling procedure 
(a two-stage, randomized stratified procedure) based on expected mortality.
In the first stage, based on the 1988 Tanzanian census, the census clusters were randomly 
selected after stratifying them according to mortality rates and agro-climatic zones. 
Households were then divided into ``high-risk'' and ``low-risk'' groups in the 
second stage, based on illness and death of households in the 12 months 
before the enumeration process.
Finally, households were randomly sampled from the groups.
For further details on the sample selection, see World Bank (2004).
}
The average interval between each of the survey round was between six and seven months, 
and interviews of households took place on a rolling basis throughout the year. 
The Kagera region is in the northwestern corner of Tanzania, west of Lake Victoria, and 
borders Uganda, Rwanda, and Burundi. 
The population of the area---1.3 million in 1988---is 
overwhelmingly rural and primarily engaged in agriculture.

% \begin{figure}[ht] 
%  \centering
%  \includegraphics[width=0.6\textwidth]{../staticPDF/kagera.pdf}
%  \caption{Map of Kagera region}
%   \label{fig:kagera_map}
% \end{figure}

@@

Table \ref{tab:desc_stat_croploss} shows the main explanatory variable of interest 
and the 5 outcomes that we examine.
The main explanatory variable is a dummy for whether the household experienced 
an accidental loss of crops, which covers any crop lost due to insects, 
rodents, fire, rotting, or other calamities, greater than TZS 200 since the 
prior survey round (or prior 1 year for the first round).
% We create a dummy for accidental crop losses greater than 200 TZS,
% which is our preferred measure of income shock.
The TZS 200 cut-off is equivalent to approximately 5th 
percentile of all household savings.
% That period is approximately 6 to 7 months for the last three surveys, but covers 
% the previous 12 months for the first survey. 
% [MAY HAVE TO EXPLAIN MORE ABOUT WHAT THIS MEANS FOR OUR ESTIMATES]
There is substantial variation in the number of households that experience a 
crop loss between the 4 survey rounds.
In the 12 months prior to the first survey round just over 70\% of households 
lost crops, whereas just over 25\% lost crops between the 1st and 2nd survey round.
That falls even further for the 3rd and 4th rounds, with less than 5\% 
reporting crop loss of more than TZS 200 for both.
Our results are robust to defining crop loss as any crop loss, with the 
coefficients slightly smaller and slightly less statistically 
significant as expected. 
% Crop losses between the valuation of 0-200 TZS are not coded as an income shock,
% because of the small magnitude (equal to  of savings). 
% and unlikely to affect fertility decisions of households. 
Furthermore, our results are also qualitatively similar if we use a log of the amount of 
crop lost or a linear specification of crop loss instead of the dummy variable.%
\footnote{
Results using these alternative specifications of crop loss are available in the Appendix.
}

@@

We estimate the relationship between household income shocks and decisions
on timing of fertility. 
To account for potential correlation between unobservable household
characteristics and both shocks and fertility decisions, we employ a fixed effect model.
Our estimates show that both the likelihood of pregnancy and birth decrease in
response to a crop loss and that contraceptive use, especially use of traditional 
methods, such as abstinence and rhythm, significantly increases.

@@

Our main conclusion is that households consciously delay 
fertility through contraceptive use to cope with income shocks. 
This is especially of interest since this is accomplished 
predominately through the use of traditional contraceptives.
If traditional contraceptives have low efficacy, a natural question is 
how the substantial reductions in likelihood of births and pregnancy 
come about.
We present suggestive evidence that show that only those households
with a strong incentive, i.e. those exposed to crop loss, are 
effective users of traditional contraceptives.%
\footnote{
This is in line with the results in \cite{Rosenzweig1989},
where women with the highest opportunity cost of children
are also the most effective user of traditional contraceptives,
although the setting is very different.
}

@@

The data unfortunately do not allow us to establish whether shocks
have long-term impacts on fertility outcomes.
Results from Guatemala indicate that only shocks that occur 
toward the end of a woman's reproductive life is likely to have an 
impact on final fertility because the ability to ``make up'' 
postponed births are smaller \citep{Portner2014}.
There is, therefore, little reason to believe that temporary shocks,
like the crop losses studied here, lead to over-all changes in fertility,
although the underlying \emph{risk} of shocks may.

[Lead with this or the next sentence?]
Our result that people are able to postpone fertility using mainly 
traditional contraceptive do, however, have implications for our 
understanding of long-term fertility.
[Alternative lead]
What implications do our results that people are able to postpone
fertility using mainly traditional contraceptive have for our
understanding of long-term fertility?
% What we do not know is whether the psychic cost of using traditional 
% contraceptives is so high that although households are willing to use 
% them to deal with short-term shocks, they become difficult to use for 
% long-run moderation of fertility.
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.%
\footnote{
See, for example, \cite{Portner2014a} and the literature surveyed therein.
}
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal clearly plays little or no role in this area.

Finally, our focus on traditional contraceptives does
not indicate that access to modern contraceptives is unimportant.
Although use of modern contraceptives is low in the area we show
they are also effective in preventing pregnancies and births after a shock.
Children born after shocks are likely to fare worse than children born 
during better times.
The mother is also likely to fare better if she does not give birth 
immediately following an economic shock that may impact her ability 
to recover and care for the new child.
Our results imply that providing better access to modern contraceptives 
during times of economic shocks may improve the ability of households 
to postpone births.
In turn, this could lead to better mother and child health outcomes.
